---
title: "Modified_portfolio_BL"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

---
title: "Tangency portfolio and Black-Litterman approach"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

# 3. Introduction

## 3.1 Packages

We use the following libraries:

1.  tidyverse: This package is a powerful toolbox that combines various
    R packages to streamline data manipulation, visualization, and
    analysis. It includes versatile tools like ggplot2 for visuals,
    dplyr for data tweaks, and tidyr for tidying data.

2.  quantmod: quantmod specializes in quantitative financial modeling
    and data analysis. It equips us with tools to handle financial data,
    perform technical analysis, and model financial time series.

3.  DataExplorer: DataExplorer aids in the initial data exploration
    phase. It provides quick insights through functions for generating
    summary statistics, visualizing missing data, and plotting variable
    distributions.

4.  corrplot: This package is your go-to for visualizing correlation
    matrices. It offers functions to create informative correlation
    plots, helping you grasp relationships between data variables.

5.  scales: scales complements R graphics, especially in conjunction
    with ggplot2. It furnishes various scales and formatting functions
    to customize plot aesthetics, including axes, labels, and color
    scales

```{r}
library(tidyverse)
library(quantmod)
library(DataExplorer)
library(corrplot)
library(scales)
```

## 3.2 Investment universe

We have carefully chosen a set of five trackers (ETFs) to represent our
univestment universe. An ETF can be defined as a financial product that
is based on a basket of different assets, to replicate the actual
performance of each selected investment. An ETF has more or less the
same proportion of the underlying components of the basket, depending on
the style of management of the asset manager.

Our selection strategy is based on several fundamental considerations:

-   Investment universe: The investment universe aims at representing
    different asset classes in order to construct a multi-asset
    portfolio.

-   Diversification strategy: Our commitment to constructing a
    well-rounded portfolio has been affirmed through diversification
    across various sectors and industries. Such diversification is a
    critical risk management tool, curbing overexposure to a single
    sector or asset class.

-   Historical data availability: The selected trackers benefit from a
    rich repository of historical data and extensive research. The
    accessibility of such data enable these ETFs to rigorous portfolio
    management analysis and comprehensive research.

The provided R code defines the tickers for these trackers, allowing you
to access the financial data. This data can then be employed to conduct
a range of analyses, such as return calculations, risk assessments, and
portfolio optimization, to test and implement various investment
strategies.

We source our data from Yahoo Finance, a highly reputable platform
renowned for its financial data. Our analysis covers a comprehensive
three-and-a-half-year period, allowing us to evaluate the performance of
our selected asset class both before and after the onset of the COVID-19
pandemic.

```{r}
# Define the stock tickers you want to fetch data for
stock_tickers <- c("VTI", "EMGF", "IEF", "DBC", "GLD")

# Fetch historical stock data from Yahoo Finance
start_date <- "2019-01-01"  # Replace with your desired start date
end_date <- "2023-08-31"    # Replace with your desired end date

# Use getSymbols to fetch data for all stock tickers
stock_data <- getSymbols(stock_tickers, from = start_date, to = end_date, auto.assign = TRUE)
```

## 3.3 Data cleaning

We initiate data cleaning, a crucial step in data analysis:

-   Extracting Adjusted Closing Prices: We use the lapply function to
    obtain adjusted closing prices, commonly used for return
    calculations, for a list of stock tickers. This code iterates
    through each ticker and fetches the adjusted closing prices.

-   Combining Stock Returns: To consolidate the adjusted closing prices,
    we employ cbind, creating a data frame where each column represents
    a specific stock's adjusted closing prices.

-   Renaming Columns: We label the columns with their corresponding
    stock tickers, facilitating easy identification.

-   Handling Missing Data: We remove rows with missing values from the
    dataset for quality and consistency.

```{r}
# Extract adjusted closing prices (which are typically used for returns)
stock_returns <- lapply(stock_tickers, function(ticker) {
  Ad(get(ticker))
})

# Combine stock returns into a single data frame
fin_prices <- do.call(cbind, stock_returns)

# Eliminate missing data in time series
fin_return <- na.omit(fin_prices)  # Remove rows with missing values

```

Arithmetic returns are used for performance assessment, offering a clear
view of gains and losses, suitable for short-term analysis, and ensuring
transparency.

```{r}
# Computation of returns
arith_fin_returns = diff(fin_return)/lag(fin_return)
head(arith_fin_returns, n=3)
arith_fin_returns <- arith_fin_returns[-1, ]


```

In order to ensure that the data is properly imported, with non missing
valiues and proper data cleaning procedure, we can implement this line
of code in order to learn more about data.

```{r}
# Data exploration for the return time series
plot_intro(arith_fin_returns)
```

We plot the rebased performance of the trackers selected in our
investment universe. From the graph below, we can clearly see the
outperformance of equity tracker (VTI) by a considerable margin with
respected to the other asset classes.

The outperformance of equity returns during the period from 2019 to 2023
can be attributed to several key factors, with a strong emphasis on the
role of equity markets, particularly the US market, as the main driver
of these returns. Here's a justification for this phenomenon:

-   Global economic growth: The period from 2019 to 2023 witnessed
    relatively strong global economic growth, with major economies
    recovering from the aftermath of the 2008 financial crisis. This
    growth provided a favorable backdrop for equity markets, as rising
    economic activity often correlates with higher corporate earnings.

-   Monetary policy support: Central banks, including the Federal
    Reserve in the United States, implemented accommodative monetary
    policies. Low-interest rates and quantitative easing measures aimed
    to stimulate economic activity and encourage investment in riskier
    assets like equities.

-   US equity dominance: The US equity market, represented by indices
    like the S&P 500, played a central role in driving global equity
    returns. The US market is home to a diverse range of sectors,
    including technology, healthcare, and consumer goods, which
    collectively contributed to robust earnings growth.

-   Technological advancements: The tech sector, dominated by US
    companies, experienced significant advancements during this period.
    Innovations in areas such as cloud computing, e-commerce, and
    digital services led to substantial revenue growth for many tech
    firms, lifting their stock prices.

-   Investor sentiment: Favorable investor sentiment, driven by
    expectations of continued economic expansion and corporate profits,
    played a crucial role. Positive sentiment often leads to increased
    capital flows into equities.

-   Sectoral performance: Certain sectors, like technology and
    healthcare, significantly outperformed others due to their
    resilience and growth prospects. The performance of these sectors,
    coupled with their weight in major indices, contributed to overall
    equity outperformance.

-   Globalization: Equity markets became more interconnected than ever,
    enabling investors to diversify globally. The interplay of global
    economic factors and cross-border investments amplified returns.

-   Risk appetite: As economic conditions improved, investors exhibited
    a greater appetite for risk, favoring equities over traditional
    safe-haven assets like bonds.

In summary, the outperformance of equity returns during the 2019-2023
period can be largely attributed to a combination of global economic
growth, accommodating monetary policies, technological advancements, and
the dominant role of the US equity market. The US market's influence,
driven by the performance of its key sectors and companies, has been a
primary driver of these strong returns.

```{r}
# Calculate Cumulative Returns
cumulative_returns <- cumprod(1 + arith_fin_returns)

# Plot Cumulative Return Performance
library(ggplot2)
autoplot(cumulative_returns, facets = NULL) +
  ggtitle("Cumulative Return Performance of the investment universe") +
  ylab("Cumulative Return") +
  xlab("Year")

```

To enhance our data analysis, we can perform statistical analysis on the
daily returns of each stock. This helps us evaluate their performance
and behavior during the observed timeframe. Please note that the figures
derived from this analysis are presented in daily returns.

```{r}
# Statistical summary of returns for each asset class

summary(arith_fin_returns)
```

In portfolio management, statistical analysis plays a key role in
understanding financial data patterns. Two essential tools are density
plots, which visualize data distributions, and QQ plots, used to assess
data conformity to theoretical distributions. These methods assist
portfolio managers in making data-driven investment decisions, enhancing
the exploration of financial data, and understanding the behavior of
selected equities' returns. As seen from the QQ plot, the results are
not normally distributed.

```{r}
# Advanced statistical analysis for the dataset

#plot_density(arith_fin_returns)
plot_qq(arith_fin_returns)
```

Correlation is a fundamental concept related to diversification, which
seeks to use poorly correlated and non correlated assets to construct a
portfolio. Diversification aims to boost returns while minimizing risk
by investing in assets that respond differently to various events.
Prudent investors seeking risk mitigation will diversify their
portfolios.

In our investment universe, opportunities for diversification arise.
Defensive assets like gold (TICKER: GLD) can effectively diversify when
combined with more aggressive ones. For instance, the Vanguard Total
Index tracker (TICKER: VTI) has a poor correlation with gold (TICKER:
GLD). iShares 7-10 years Treasuries Index tracker (TICKER: IEF) exhibits
negative correlation with respect to international and emerging market
equities.

```{r}
# Correlation analysis

library(corrplot)
corrplot( cor(arith_fin_returns), type='lower', method = "shade", tl.col = 'black', cl.pos = "r", tl.cex = 1)
```

# 4. Modelling of the portfolio

## 4.1 Implementing the Tangency portfolio (TP)

In time series data analysis, a crucial application involves splitting
the dataset into training and testing sets based on temporal order. This
is vital when working with data that evolves over time, such as stock
prices or weather observations.

The training set consists of historical data, while the testing set
contains more recent observations. This division allows analysts to
assess predictive models and forecasting techniques. By training the
model on historical data and evaluating it with recent data, we can
determine its ability to make accurate predictions. This is especially
important in finance for understanding and forecasting trends over time.

For modeling purposes, we'll create two sub-samples with defined
parameters.

```{r}
# Specify the number of rows for the first subsample
num_rows_subsample1 <- 400

# Calculate the total number of rows in your dataset
total_rows <- nrow(arith_fin_returns)
```

1.  The first sub-sample will cover the first 400 trading days covered
    in the data set.

```{r}
# Create the first subsample
subsample1 <- arith_fin_returns[1:num_rows_subsample1, ]
plot_intro(subsample1)

```

2.  The second sub-sample will cover the rest of the data set, covering
    the equivalent of 773 trading days.

```{r}
# Create the second subsample by excluding the rows in the first subsample
subsample2 <- arith_fin_returns[(num_rows_subsample1 + 1):total_rows, ]
plot_intro(subsample2)
```

We implement the parameters of the model in order to compute the
Tangency Portfolio:

```{r}
n <- ncol(subsample1)
T <- nrow(subsample1)
e <- rep(1, n)
perio <- 252
rf <- 0

```

The parameter sigma, representing the unbiased covariance matrix
(similar to the procedure implemented when computing the GMV portfolio),
can be computed using the following procedure:

```{r}
mu <- colMeans(subsample1) * perio - rf
Sigma <- cov(subsample1) * (T - 1) / (T - n - 2) * perio
A <- t(e) %*% solve(Sigma) %*% mu
omega <- 1 / as.numeric(A) * solve(Sigma) %*% mu

returns = mu %*% omega
```

The portfolio allocation consists of both long positions, where
investors expect favorable returns, and short positions, where they
anticipate weaker performance. Here's a more detailed explanation of
each asset class allocation:

Long Positions:

-   International Equities (Ticker: VTI, Allocation: 25.38%): This
    segment of the portfolio is dedicated to international equities,
    with VTI making up a substantial portion at 25.38%. Investors are
    optimistic about the performance of international equities, which is
    why they have allocated a significant portion of their capital to
    this asset class.

-   US Short-Term Maturity Bond (Ticker: IEF, Allocation: 77.03%): The
    largest allocation in the long positions is in US short-term
    maturity bonds, represented by IEF. This allocation, at 77.03%,
    indicates a strong conviction in the stability and income potential
    of short-term bonds.

-   Gold (Ticker: GLD, Allocation: 15.27%): The allocation to gold,
    represented by GLD, has decreased to 15.27%. Gold is often
    considered a safe-haven asset, and this allocation suggests that
    investors still see value in holding gold as a hedge or store of
    value, albeit at a slightly reduced allocation.

Short Positions:

-   Emerging Markets (Ticker: EMGF, Allocation: -6.41%): The short
    segment of the portfolio is allocated to assets where weak
    performance is expected. In the case of emerging markets, the
    allocation is bearish at -6.41%. This indicates that investors are
    taking a negative stance on the performance of emerging market
    assets, expecting their values to decline.

-   Commodities (Ticker: DBC, Allocation: -11.26%): The allocation to
    commodities, represented by DBC, is even more conservative at
    -11.26%. This suggests a strongly bearish outlook on commodities,
    where investors anticipate a significant decrease in their value.
    This allocation reflects a cautious approach towards commodities in
    the portfolio.

In summary, the portfolio is strategically balanced between long
positions, where investors have confidence in the assets' growth
potential, and short positions, where they are expecting weaker
performance. The specific allocations provide insights into the level of
conviction and sentiment regarding each asset class.

```{r}
# Plot of Tangency portfolio

barnames <- c("VTI", "EMGF", "IEF", "DBC", "GLD")
barplot(
  as.numeric(omega),
  col = 'Black',
  names.arg = barnames,
  ylim = c(-0.5, 1)
)
```

## 4.2 Implementing Black-Litterman approach

<!-- As described in their investment outlook (BlackRock, 2023), we can base -->

<!-- our market outlook on the following analysis conducted by BlackRock, one -->

<!-- of the best asset management firm worldwide. The reasoning of each -->

<!-- outlook will help construct a assumptions on the value of our views in -->

<!-- portfolio allocation under the Black-Litterman framework. -->

<!-- -   International equities: BlackRock is staying neutral on -->

<!--     international equities. Valuations in many international markets are -->

<!--     relatively low, and there are indications that earnings are on the -->

<!--     rise. Nevertheless, investors should be mindful of the risks -->

<!--     associated with international investments, such as currency -->

<!--     fluctuations and political instability. -->

<!-- -   Emerging markets: BlackRock is taking a neutral stance on emerging -->

<!--     markets. While emerging markets offer the potential for attractive -->

<!--     returns, they also come with higher risk compared to developed -->

<!--     markets. Therefore, investors should carefully assess their risk -->

<!--     tolerance and investment objectives before venturing into emerging -->

<!--     markets. -->

<!-- -   Short-term US Treasuries: BlackRock prefers short-term US -->

<!--     Treasuries. The outlook for short-term US Treasuries is positive, as -->

<!--     the Federal Reserve is likely to keep increasing interest rates to -->

<!--     combat inflation. Nevertheless, investors should be aware that the -->

<!--     Fed might raise rates more aggressively than anticipated, leading to -->

<!--     potential losses in short-term US Treasuries. -->

<!-- -   Commodities: BlackRock is favoring commodities with a higher -->

<!--     allocation. Although commodity prices have been fluctuating in -->

<!--     recent months, they are expected to stay strong due to robust demand -->

<!--     from China and the ongoing conflict in Ukraine. However, investors -->

<!--     should note that commodity prices tend to follow cyclical patterns, -->

<!--     and they may not continue rising indefinitely. -->

<!-- -   Gold: BlackRock has a higher allocation to gold in its portfolios. -->

<!--     The short-term prospects for gold look good because investors are -->

<!--     turning to safe-haven assets due to concerns about global economic -->

<!--     growth and geopolitical tensions. However, in the long run, gold -->

<!--     prices are likely to be influenced more by changes in real interest -->

<!--     rates. -->

In order to construct our views matrix, we can base our assumlptions on
the economic outlooks provided from a group of asset managers to
confront outlooks and assess their perspective on markets. We will delve
deeper into the outlooks from BlackRock, Goldman Sachs Asset Management
(GSAM), and J.P. Morgan Asset Management (JPMAM) on gold, international
equities, commodities, short-term US Treasuries, and emerging markets:

-   **International equities:** BlackRock and GSAM are neutral on
    international equities, while JPMAM is overweight. This means that
    BlackRock and GSAM are neither optimistic nor pessimistic about the
    outlook for international equities, while JPMAM is bullish.
    International equities have underperformed US equities in recent
    years, but they are now starting to look more attractive. Valuations
    in many international markets are relatively low, and there are
    signs that earnings growth is picking up. However, investors should
    be aware of the risks associated with investing in international
    markets, such as currency fluctuations and political instability.

-   **Emerging markets:** BlackRock and JPMAM are neutral on emerging
    markets, while GSAM is underweight. This means that BlackRock and
    JPMAM are neither optimistic nor pessimistic about the outlook for
    emerging markets, while GSAM is bearish. Emerging markets offer the
    potential for attractive returns, but they are also more risky than
    developed markets. Investors should carefully consider their risk
    tolerance and investment goals before investing in emerging markets.

-   **Gold:** All three asset managers are overweight gold, which means
    that they believe that gold is a good investment at this time. Gold
    is often seen as a safe haven asset, and it is often sought after by
    investors during times of economic uncertainty. The current economic
    outlook is uncertain, and inflation is high, which is likely to
    support gold prices in the near term.

-   **Commodities:** All three asset managers are overweight
    commodities, which means that they believe that commodities are a
    good investment at this time. Commodity prices have been volatile in
    recent months, but they are likely to remain supported by strong
    demand from China and the ongoing war in Ukraine. However, investors
    should be aware that commodity prices can be cyclical, and they may
    not continue to rise indefinitely.

-   **Short-term US Treasuries:** BlackRock and GSAM are overweight
    short-term US Treasuries, while JPMAM is neutral. This means that
    BlackRock and GSAM are bullish on short-term US Treasuries, while
    JPMAM is neither optimistic nor pessimistic. Short-term US
    Treasuries are a good option for investors who are looking to
    preserve capital and earn some income. The Federal Reserve is
    expected to continue raising interest rates in the near term, which
    will push up yields on short-term US Treasuries.

Overall, the table shows that there is a general convergence of outlooks
on gold, commodities, and short-term US Treasuries. All three asset
managers are overweight these asset classes. However, there is some
divergence of outlooks on international equities and emerging markets.
BlackRock and GSAM are neutral on international equities, while JPMAM is
overweight. BlackRock and JPMAM are neutral on emerging markets, while
GSAM is underweight.

In summary, our combined views suggest the following outlook scenario
for each asset class:

-   **International equities** (TICKER: VTI): Strongly overweight

-   **Emerging Markets** (TICKER: EMGF): Relatively underweight

-   **US short term maturity bond** (TICKER: IEF): Strongly overweight

-   **Commodities** (TICKER: DBC): Relatively underweight

-   **Gold** (TICKER: GLD): Relatively overweight

```{r}
Q <- numeric(5)
Q[1] <- 0.00 # neutral view for US equity index (VTI)
Q[2] <- 0.00 # neutral view for Global bond index (EMGF)
Q[3] <- 0.05 # positive view for global equity index (IEF)
Q[4] <- 0.1 # positive view for commodity index (DBC)
Q[5] <- 0.08 # postive view for the gold index (GLD)
tau <- 0.95
```

The Black-Litterman optimization approach is a widely-used method for
improving the estimation of expected returns and making asset allocation
decisions. It addresses the limitations of traditional mean-variance
optimization by incorporating investor views and market equilibrium
considerations. The following key steps are performed:

Mixed estimation of returns: The code starts by computing expected
returns using mixed estimation. This means combining the market
equilibrium-based returns (mu) with the views-based returns (Q) using a
blending process. The parameter tau adjusts the strength of the
blending. This approach allows investors to incorporate their subjective
views on assets into the asset allocation process while considering the
overall market dynamics.

Tactical allocation with views directly: This part of the code
calculates the tactical asset allocation weights based on views
directly. It uses a matrix A_Q and the matrix of uncertainty in views
(omega_Q) to determine the optimal portfolio weights. These weights
reflect the investor's views on specific assets and are adjusted
according to their confidence in those views.

Tactical allocation with mixed estimation: Similarly, this part
calculates tactical asset allocation weights, but this time, it uses the
mixed estimation approach. It calculates A_mixed and omega_mixed based
on the mixed estimation of expected returns. This allows investors to
take a balanced approach, combining both market equilibrium
considerations and their own views in the asset allocation process.

Black-Litterman offers a flexible framework for investors to incorporate
their insights and views while maintaining a connection to market
equilibrium. It's a valuable tool for optimizing portfolio allocation
decisions, especially when investors have specific expectations or
insights about the market.

Source [1] Black, F., & Litterman, R. B. (1992). Global portfolio
optimization. Financial Analysts Journal, 48(5), 28-43.

```{r}

#Mixed estimation of returns

Omega <- diag(diag(Sigma), n, n)
mu_mixed <-
  solve(solve(tau * Sigma) + solve(Omega)) %*% (solve(tau * Sigma) %*% mu +
                                                  solve(Omega) %*% Q)

#Tactical allocation with views directly

A_Q <- t(e) %*% solve(Sigma) %*% Q
omega_Q <- 1 / as.numeric(A_Q) * solve(Sigma) %*% Q
#barplot(as.numeric(omega_Q), col = 'black', names.arg = barnames, ylim = c(-0.5,1.5))


```

We can add the following comments on the new portfolio allocation under
Black-Litterman approach, including asset views.

**Asset Class Allocations**:

-   **International Equities (Ticker: VTI, Allocation: 28.20%)**: The
    allocation to international equities is 28.20%. This allocation
    represents a "strongly overweight" position compared to the neutral
    stance of the BlackRock outlook, indicating a high level of
    confidence in the favorable performance of international equities.
    However, it's essential to note that there are differing views among
    asset managers. BlackRock and GSAM are neutral on international
    equities, neither optimistic nor pessimistic about their outlook,
    whereas JPMAM is "overweight," indicating a bullish perspective.
    This divergence in views reflects the varying assessments of
    international equity performance.

-   **Emerging Markets (Ticker: EMGF, Allocation: -11.28%)**: The
    allocation to emerging markets is -11.28%, indicating a "relatively
    underweight" position. This aligns with the BlackRock outlook, which
    takes a neutral stance on emerging markets. JPMAM shares this
    neutral view, while GSAM adopts a more cautious stance by being
    "underweight," indicating a bearish outlook. The divergence in views
    highlights the risk-return trade-off associated with emerging
    markets.

-   **US Short Term Maturity Bond (Ticker: IEF, Allocation: 80.80%)**:
    The allocation to US short-term maturity bonds is 80.80%, reflecting
    a "strongly overweight" position. This significant overweight
    allocation aligns with the views of both BlackRock and GSAM, which
    are bullish on short-term US Treasuries. In contrast, JPMAM
    maintains a "neutral" position, neither optimistic nor pessimistic.
    Short-term US Treasuries are seen as a favorable option for capital
    preservation and income generation, given the expected rise in
    yields.

-   **Commodities (Ticker: DBC, Allocation: -6.91%)**: The allocation to
    commodities is -6.91%, indicating a "relatively underweight"
    position. This allocation is consistent with the outlook of all
    three asset managers, which are "overweight" commodities, signifying
    a positive investment view. While commodity prices have exhibited
    volatility, they are expected to remain supported by strong demand
    from China and geopolitical tensions. Nevertheless, investors should
    be aware of the cyclical nature of commodity prices, which may not
    experience continuous growth.

-   **Gold (Ticker: GLD, Allocation: 9.19%)**: The allocation to gold is
    9.19%, reflecting a "relatively overweight" position. This aligns
    with the outlook of all three asset managers, who are collectively
    "overweight" on gold. This indicates a shared belief that gold is a
    sound investment at this time, particularly as it is considered a
    safe-haven asset sought after during periods of economic
    uncertainty. Given the current economic outlook's uncertainty and
    high inflation, gold prices are expected to be supported in the near
    term.

**Comments on Portfolio Allocation**:

-   The portfolio's strong overweight position in international equities
    signifies a bullish stance, despite differing views among asset
    managers. BlackRock and GSAM maintain a neutral position, while
    JPMAM's overweight perspective suggests a higher degree of
    confidence in the outlook for international equities.

-   The relatively underweight position in emerging markets aligns with
    the cautious approach recommended by BlackRock. Differing views from
    GSAM and JPMAM indicate varying risk appetites, acknowledging the
    challenges associated with investing in these markets.

-   The significant overweight allocation to US short-term maturity
    bonds is a key feature of the portfolio allocation. This reflects
    confidence in the short-term Treasuries and is supported by
    BlackRock and GSAM's bullish stance, with JPMAM taking a neutral
    position.

-   The relatively underweight allocation to commodities is consistent
    with the cyclical nature of these assets. The portfolio takes a more
    conservative stance, despite expectations of continued strength in
    commodity prices.

-   The relatively overweight position in gold underscores its potential
    as a safe-haven asset in the short term, aligning with the views of
    all three asset managers.

Overall, the portfolio allocation under the Black-Litterman approach
takes into account a broader spectrum of views from different asset
managers and integrates them into a more specific allocation that
reflects varying degrees of optimism and caution across different asset
classes. Incorporating views into the portfolio has resulted in diverse
adjustments across asset classes, reflecting the investor's evolving
outlook on each. These changes are indicative of a dynamic strategy that
aims to capitalize on various market scenarios. It's crucial to consider
these allocations within the context of the investor's overall
investment objectives and risk tolerance, as they play a pivotal role in
shaping the portfolio's risk-return profile.

```{r}
#Tactical allocation with mixed estimation
A_mixed <- t(e) %*% solve(Sigma) %*% mu_mixed
omega_mixed <- 1 / as.numeric(A_mixed) * solve(Sigma) %*% mu_mixed
barplot(as.numeric(omega_mixed), col = 'black', names.arg = barnames, ylim = c(-1,1))
```
